Analysing Constraint Grammars with a SAT-solver

### 2016-03-08 paper/lrec/eval_futurework.tex
more numbers and stuff on spanish

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# true positives  & 7             & n         & m    \\ \hline
\# false positives & 0             & n         & m    \\ \hline
\clock{} with amb. 
           classes & ??s           & + 1h      & 30m-1h   \\ \hline
% \clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} no amb. 
           classes & 10ish sec       & ??min ??s    & ?h ??min    \\ \hline

@@

We include two versions of the Spanish grammar in table~\ref{table:res}: in column 2, we added the lemmas and word forms as described in Section~\ref{sec:realistic_readings}, and in column 3, we just added each word form and lemma as individual readings, allowed to combine with any other reading. \todo{check if true: This did not affect the amount of conflicts detected, but we found some differences in the generated output; examples will be shown later in this section.}

Again, we can classify the conflicts into internal and interaction. As an example of internal conflict, there are \todo{n} rules that use \texttt{SET Cog = (np cog)}---the problem is that the tag \texttt{cog} does not exist in the Apertium dictionary for Spanish. It is likely that this grammar has been written for an earlier version, where such tag has been in place.

The Spanish grammar has a high number of interaction conflicts, \todo{n} with the same pattern:

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE (a) IF (-1 x) (1 y) ;
REMOVE (a) IF (-1 x) (1 y) (NOT -1 z);
\end{verbatim}
\end{itemize}

This is another instance of more general rule applying before more
specific. Negated context works as well to make a rule stricter. The
second rule will never apply, because the first will catch all
cases. For a full list of found conflicts, see \todo{\url{actual://link.here/spanish/error.log}}

@@

The second challenge comes from the larger size of the grammar. Whereas the Spanish and Dutch had only tens of word forms or lemmas, the Finnish grammar specifies around 900 of them.
Due to both of these factors, the procedure described in Section~\ref{sec:realistic_readings} would have exploded the number of readings, so we simply took the lemmas and word forms, and added them as single readings. 
In cases where they were combined with another tag in the grammar, we took that reading directly: for instance, we took both \emph{aika} and ``\emph{aika} n'' from \texttt{SELECT "aika" + N},
but nothing from \texttt{SELECT Pron + Sg}. This method gave us 1588 additional readings.

The third were not able to create ambiguity class constraints---expanding the Finnish morphological lexicon results in \todo{100s of gigabytes?} of word forms, which is simply too big for our method to work. 
For future development, we will see if it is possible to manipulate the finite automata directly, instead of relying on the output in text.

@@

\begin{itemize}
\item[] \todo{Put back the rule with 0 gen and 0C nom!!! that was actually nice and legit}

\item[]\texttt{SELECT ADV (-1C AD) (NOT 0 PP) (NOT 1 A) ;}
\end{itemize}

This is an internal conflict, which is not trivial to see: the problem is that the definition of \texttt{PP} includes \texttt{adv} among others. With the sets expanded, this rule becomes \texttt{SELECT adv IF (NOT 0 pp|adv|adp|po|pr) ...}.

@@

\begin{itemize}
\item[]\begin{verbatim}
SELECT:r1 .. IF (-1 Comma/N/Pron/Q) ;
SELECT:r2 .. IF (-2 ..) (-1 Comma) ;
\end{verbatim}
\end{itemize}

The rules above have been simplified to show only the relevant part.
The conflict lies in the fact that \texttt{Comma} is a subset of \texttt{Comma/N/Pron/Q}:
there is no way to trigger $r2$ without placing a comma in position -1, and thereby triggering $r1$.

@@

\subsection{Performance} 
The running time of the grammars goes from seconds to hours. 
As can be seen from the times, increasing the number of readings
matters more than increasing the number of rules.
In addition, the ambiguity class constraints increase the time.

@@

>>>>

\hline
                   & \textsc{nld}  & \textsc{spa}lem  & \textsc{spa}nolem \\ \hline
\# rules           & 59            & 279       & 279     \\ \hline
\# readings        & 336           & 3905      & 1735    \\ \hline
\# true positives  & 7             & n         & m    \\ \hline
\# internal/interaction & 6 + 1    & n         & + 25    \\ \hline

\# false positives & 0             & n         & 5    \\ \hline
\# internal/interaction & 0        & n         & 5 +   \\ \hline

\clock{} with amb. 
           classes & ??s           & 5h 42m    &  11m25  \\ \hline
% \clock{} last rule & ??s           & ??s       & ?min ??s    \\ \hline
\clock{} no amb. 
           classes & 10ish sec       & 14m 16s    & ?h ??min    \\ \hline

@@

We include two versions of the Spanish grammar in table~\ref{table:res}: in column 2, we added the lemmas and word forms as described in Section~\ref{sec:realistic_readings}, and in column 3, we just added each word form and lemma as individual readings, allowed to combine with any other reading. 
This latter version ran much faster, but also resulted in 5 false positives, all for rules that require a lemma in a condition.
In addition, there were differences in the generated output; examples will be shown later in this section.

Again, we can classify the conflicts into internal and interaction. As an example of internal conflict, there are two rules that use \texttt{SET Cog = (np cog)}---the problem is that the tag \texttt{cog} does not exist in the Apertium dictionary for Spanish. As another example, four rules require a context word tagged as NP with explicit number, but the lexicon does not indicate any number with NPs. It is likely that this grammar has been written for an earlier version, where such tags have been in place.

As for the 25 interaction conflicts, there were only 9 distinct rules that rendered 25 other rules ineffective.
In fact, we can reduce these 9 rules further into 3 different groups: 4 + 4 + 1, where the groups of 4 rules are variants of otherwise identical rule, with  combinations feminine singular, masculine singular, feminine plural and masculine plular.
An example of such conflict is below (gender and number omitted for readability):

\begin{itemize}
\item[] 
\begin{verbatim}
# NOM ADJ ADJ
SELECT A OR PP IF (-2 N) 
 (-1 Adj_PP) (0 Adj_PP) (NOT 0 Det);

# NOM ADJ ADJ ADJ
SELECT A OR PP IF (-3 N) (-2 N) 
 (-1 Adj_PP) (0 Adj_PP) (NOT 0 Det);
\end{verbatim}
\end{itemize}

For a full list of found conflicts, see \todo{\url{actual://link.here/spanish/error.log}}

@@

The second challenge comes from the larger size of the grammar. Whereas the Spanish and Dutch had only tens of word forms or lemmas, the Finnish grammar specifies around 900 of them.
Due to both of these factors, the procedure described in Section~\ref{sec:realistic_readings} would have exploded the number of readings, so we simply took the lemmas and word forms, and added them as single readings. 
In cases where they were combined with another tag in the grammar, we took that reading directly: for instance, we included both \emph{aika} and ``\emph{aika} n'' from the rule \texttt{SELECT "aika" + N},
but nothing from the rule \texttt{SELECT Pron + Sg}. This method gave us 1588 additional readings.

Finally, we were not able to create ambiguity class constraints---expanding the Finnish morphological lexicon results in \todo{100s of gigabytes?} of word forms, which is simply too big for our method to work. 
For future development, we will see if it is possible to manipulate the finite automata directly, instead of relying on the output in text.

@@

\begin{itemize}
\item[] \item[]\texttt{SELECT "oma" + Gen IF .. (0C Nom) ;}
\item[]\texttt{SELECT ADV (-1C AD) (NOT 0 PP) (NOT 1 A) ;}
\end{itemize}

Both of these are internal conflicts, which may not be trivial to see: 
the first rule requires the target to be genitive and unambiguously nominative: 
the problem in the first rule is that the target is required to be  the problem is that the definition of \texttt{PP} includes \texttt{adv} among others. With the sets expanded, this rule becomes \texttt{SELECT adv IF (NOT 0 pp|adv|adp|po|pr) ...}.

@@

\begin{itemize}
\item[]\begin{verbatim}
SELECT .. IF (-1 Comma/N/Pron/Q) ;
SELECT .. IF (-2 ..) (-1 Comma) ;
\end{verbatim}
\end{itemize}

The rules above have been simplified to show only the relevant part.
The conflict lies in the fact that \texttt{Comma} is a subset of \texttt{Comma/N/Pron/Q}:
there is no way to trigger the second without placing a comma in position -1, and thereby triggering the first rule.

@@

\subsection{Performance} 
The running time of the grammars ranges from seconds to hours. 
As can be seen from the times, increasing the number of readings
matters more than increasing the number of rules.
In addition, the ambiguity class constraints increase the time.

@@

%Spanish lemmahack: time to create the formulas 5m32.214s

%Spanish no lemmahack: time to create the formulas  62m6.049s
